---
title: "Market Access Somalia"
author: "Valentin Böhmer"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float:
      collapsed: true
    theme: cerulean
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))

options(openrouteservice.url = "http://0.0.0.0:8080/ors")

```

# Intro

Analysis of market access in Somalia. 
We will look at food market locations in Somalia and assess their accessibilty by regular sampled points.

# Setup

This section for reproducibility.
We use R in the version 4.2.2.
In order to be able to reproduce this analysis you need an openrouteservice API key.
Get one [here](https://openrouteservice.org/sign-up/).
Put it in a text file file called `config.txt` in the working directory.
It has no structure, just the key in the first line.

``` txt
ZXCVBNMLKJHGFDSAQWERTYUIOP
```

## libraries

Almost all of the libraries can be installed via CRAN with the command `install.packages('packagename')`.
The openrouteservice package is not on CRAN and needs to be installed via github.
With the command `remotes::install_github("GIScience/openrouteservice-r");` Also tmap in its most recent version 4 is not yet available on CRAN so we need to download and install it via github too.

```{r import libs, message=F, warning=F}
# main libraries
library(tidyverse)
library(glue)
library(sf)
library(units)
library(ggplot2)
library(tictoc)
library(openrouteservice) # remotes::install_github("GIScience/openrouteservice-r");
library(jsonlite)
library(terra)
library(exactextractr)
library(tmap) # remotes::install_github("r-tmap/tmap")
library(leaflet)
library(furrr)
library(purrr)
library(kableExtra)
library(mapsf)
library(RColorBrewer)
library(geonode4R)
library(scales)


#config <- readLines("config.txt")
#api_key <- config[1]
#user <- config[2]
#password <- config[3]
api_key <- ""
# Function to  adjust a list of coordinates to the location of the closest road segment via the snapping endpoint of ORS. ORS itself only allows for a maximum distance of 350m to snap.
ors_snap <- function(x, rowids, local=F) {
  x=coord_list
  rowids=grid5km$rowid
  local = T
  
  library(httr)
  
  # Define the body
  body <- list(
    locations = x,
    radius = 100000000 # apparently the maximum snapping distance is 5km
  )
  
  # Define the headers
  headers <- c(
    'Accept' = 'application/json, application/geo+json, application/gpx+xml, img/png; charset=utf-8',
    'Authorization' = api_key,
    'Content-Type' = 'application/json; charset=utf-8'
  )
  
  endpoint <- if(local){'http://0.0.0.0:8080/ors/v2/snap/driving-car'} else {'https://api.openrouteservice.org/v2/snap/driving-car'}
  # Make the POST request
  response <- POST( endpoint,
                    #'https://api.openrouteservice.org/v2/snap/driving-car', 
                    body = body, 
                    add_headers(headers), 
                    encode = "json")
  
  resp_content <- content(response)
  
  extract_data <- function(x, index) {
    if (is.null(x)) {
      return(data.frame(
        rowid = index,
        lon = NA,
        lat = NA,
        snapped_distance = NA
      ))
    } else {
      return(data.frame(
        rowid = index,
        lon = x$location[[1]],
        lat = x$location[[2]],
        snapped_distance = x$snapped_distance
      ))
    }
  }
  
  # Extract data from the list and create a dataframe
  df <- do.call(rbind, lapply(seq_along(resp_content$locations), function(i) extract_data(resp_content$locations[[i]], i)))
  df$rowid <- rowids
  # Convert the dataframe to an sf object
  sf_df <- df |> 
    drop_na(lon, lat) |>  # Drop rows with NA coordinates
    st_as_sf(coords = c("lon", "lat"), crs = 4326) |>  # Define the coordinate columns and set CRS
    st_sf()
  
  return(sf_df)
  
}

```

# Processing pipeline

The workflow is as follows:

* prepare market location data
* generate hexagonal 5km grid in Somalia
* add worldpop data per grid cell
* snap the grid centroids to the nearest road segment
* create a matrix for the distance and time for each market to the snapped centroids
* detect the closest market for each grid cell

By snapping we refer to the process of changing the location of an origin or destination of a route to the nearest road segment. This is necessary because the openrouteservice API only allows for a maximum distance of 350m to snap. However in the context of Somalia greater snapping distances are desired as the road density in some regions might be low.

## Data input

We use the following datasets as inputs:

* pop: WorldPop population counts constrained
* market locations data
* admin boundaries level 2

```{r input, message=F, warning=F}
pop <- rast('data/som_ppp_2020_UNadj_constrained.tif')
markets_som <- read_csv("data/wfp_food_prices_som.csv")
admin <- st_read('data/Som_Admbnda_Adm2_UNDP.shp', quiet = T)
```

## Markets preprocessing

In order to use the market locations we need to create a sf object by converting the latitude and longitude columns to a point geometry column. By checking the distinct locations of the geometry column we got 44 food markets in Somalia.

```{r markets prep, message=F, warning=F, cache=T}
# Clean the latitude and longitude columns
markets_som_clean <- markets_som |> 
  filter(!is.na(latitude) & !is.na(longitude)) |> 
  slice(-1) |> 
  mutate(latitude = as.numeric(latitude),
         longitude = as.numeric(longitude),
         date = as.Date(date, format = "%Y-%m-%d"))
         
#add year
markets_som_clean <- markets_som_clean |> 
  mutate(year = as.numeric(format(date, "%Y")))

# Convert to sf object
markets_som_sf <- markets_som_clean |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 

#unique markets
markets_som_asc_sf <- markets_som_sf |> 
  distinct(geometry, .keep_all = TRUE) 

#unique markets descending
markets_som_desc_sf <- markets_som_sf |> 
  arrange(desc(row_number())) |>    # Sort the dataframe such that the last rows come first
  distinct(geometry, .keep_all = TRUE)

#combine both
markets_som_sf <- markets_som_desc_sf |> left_join(markets_som_asc_sf |> st_drop_geometry() |> select(market, year), by ="market")

#prepare final market sf
markets_som_sf <- markets_som_sf |> rename(first_year = year.y, last_year = year.x) |> 
  select(-date) |>
  mutate(rowid = row_number())
```

# Grid sampling

Finding the shortest/fastest way between a number of origins and destinations is a job for the Matrix endpoint of ors. 
Lets start with the grid sampling. We decide for a 5000m width hexagonal grid. A hexagonal grid is closer to a circle and therefore the better joice for our analysis. Within each grid cell we will use the centroid and search for road network locations to snap to.

```{r create grid, cache=T, message=F, warning=F, time_it=T}

# create 5km grid
grid5km <-
  st_make_grid(admin |> st_transform(20538), cellsize = 5000, square = F, flat_topped = T)
# convert to sf df
grid5km <- grid5km |> st_as_sf()
names(grid5km) <- 'geometry'
st_geometry(grid5km) <- 'geometry'
# get rid of all cells that are not within somalia
grid5km <-
  grid5km |> st_transform(4326) |> st_filter(admin |> st_union())
# join admin codes
grid5km <-
  grid5km |> st_join(admin |> select(admin2Name, admin2Pcod, admin1Name, admin1Pcod, admin0Name, admin0Pcod),
                     left = T,
                     largest = T)
# join pop information
grid5km$pop <- exact_extract(pop, grid5km, 'sum', progress = F)

# add rowid
grid5km <- grid5km |> mutate(rowid = row_number())

#filter for pop > 0
grid5km <- grid5km |> filter(pop > 0)
```

## Snapping

The grid is created, now snap the centroids.

```{r snap, cache=T, time_it=T}
# convert
coord_list <-
  lapply(grid5km$geometry |> st_centroid(), function(x)
    c(st_coordinates(x)[, 1], st_coordinates(x)[, 2]))

# snap the list of cords
snapped_coords <- ors_snap(coord_list,rowids=grid5km$rowid, local = T)

# join back
grid5km_snapped <- grid5km |>
  left_join(snapped_coords |> tibble(), by = "rowid") |>
  mutate(centroid = st_centroid(geometry.x))

# refactor names in order to keep, the grid geometry, centorid and snapped centroid
names(grid5km_snapped)[10:11] <- c('geometry', 'snapped_centroid')
st_geometry(grid5km_snapped) <- 'geometry'

grid5km_snapped <- grid5km_snapped |> mutate(snapped = ifelse(is.na(snapped_distance), F, T)) 

#filter for snapped values
grid5km_snapped_NA <- grid5km_snapped |> filter(snapped == F)
grid5km_snapped_notNA <- grid5km_snapped |> filter(snapped == T)
```

Awesome we now got 17738 snapped and 3045 not snapped grid cells.

## ORS matrix request and post processing

In the next chunk we will calculate the travel times and distances from each grid cell to each market. We will use the openrouteservice API for this.

```{r, echo=F}
# create a list of coordinates for the grid again but from snapped coords
coord_list <-
  lapply(grid5km_snapped_notNA$snapped_centroid, function(x)
    c(st_coordinates(x)[, 1], st_coordinates(x)[, 2]))

# run through all markets and calculate travel times an distances to each grid cell.
# For 44 markets and 17738 grid cells this is 17738*44 = 780472 results.
for ( i in 1:nrow(markets_som_sf)) { 
  # log the current iteration
  tic(glue("run {i} / {nrow(markets_som_sf)}"))
  # extract single market by iteration index
  markets_subset <- markets_som_sf[i,]
  # get the coordinates as vector
  markets_coords <-
    lapply(markets_subset$geometry, function(x)
      c(st_coordinates(x)[, 1], st_coordinates(x)[, 2]))  
  # add the market coordinates at the first position before our grid centroid coordinates
  coord_list_run <- c(markets_coords, coord_list)
  
  # run the ors matrix request
  res1 = ors_matrix(
    coord_list_run, # list of all grid and one single market coordinate at the top
    sources = 0, # index of the coordiante which shall serve as origin. In our case the market. If we dont set it we get the matrix for all coords * all cords
    metrics = c("duration", "distance"), # we want duration and distance, not only one of it
    units = "km", # units is km
    api_key = api_key, # api key
    output = 'parsed') # output as parsed json
  
  rm(coord_list_run) # get rid of the list of coordinates as we create a new one the next iteration with a different market at the top
  
  # create a dataframe from the results if it is the first iteration, otherwise create and append to the existing one
  if (i == 1) {
    result_matrix <- data.frame(
      market_id = markets_subset$rowid |> rep((res1$distances |> as.vector())[-1] |>
                                             length()),
      grid_id = grid5km_snapped_notNA$rowid,
      durations = (res1$durations |> as.vector())[-1] |> as.numeric(),
      distances = (res1$distances |> as.vector())[-1] |> as.numeric()
    ) |> tibble()
  } else {
    result_matrix <- rbind(result_matrix,data.frame(
      market_id = markets_subset$rowid |> rep((res1$distances |> as.vector())[-1] |>
                                             length()),
      grid_id = grid5km_snapped_notNA$rowid,
      durations = (res1$durations |> as.vector())[-1] |> as.numeric(),
      distances = (res1$distances |> as.vector())[-1] |> as.numeric()
    ))
  }
  toc()
}

```

Awesome we now have the distances from all 44 markets to all grids. The resulting dataframe has 780472 rows. Now we want to find the market that is closest from every grid cell in terms of traveltime.


```{r, echo=F}
# Identify the lowest duration from each grid city combination
lowest_duration <- result_matrix |> 
  group_by(grid_id) |> 
  slice_min(durations, with_ties = F) |> 
  select(grid_id, market_id, durations, distances) |> 
  rename(shortest_duration = durations)

# Join the lowest duration to the grid data
grid5km_snapped_notNA <- grid5km_snapped_notNA |>
  left_join(lowest_duration, by = c('rowid' = 'grid_id')) |>
  mutate(shortest_duration = shortest_duration / 3600) # convert seconds to hours
names(grid5km_snapped_notNA)[c(12, 13)] <-
  c('duration_h', 'distance_km') # account for naming 

# join back in the not snapped grid cells. Those will bear a NA in the shortest_duration column
grid_final <- bind_rows(grid5km_snapped_notNA,
                        grid5km_snapped_NA)
```

# Visualization and Assessment

In the following chunks some visualizations are executed to assess the results.

## Non Spatial

What is the relation between Traveltime and population for each grid cell?

```{r, echo=F, warning=F}
grid_final |> 
  ggplot(aes(x = duration_h, y = pop)) +
  geom_point() + theme_minimal() + 
  xlab('Traveltime in h') + ylab('Population')  +
  scale_y_continuous(labels = comma)
```


```{r, echo=F, warning=F}
grid_final |> 
  ggplot(aes(x = duration_h, y = pop |> log())) +
  geom_point() + theme_minimal() + 
  xlab('Traveltime in h') + ylab('Population log scaled')
```


## Spatial

Maps of population distribution, travel time to the closest market and the closest market with respective administrative boundaries and the markets locations.

### Population distribution

```{r, echo=F, warning=F}
tmap_mode('view')
tm_shape(grid_final, name = 'Population') +
  tm_polygons(
    'pop',
    fill.scale = tm_scale_intervals(
      values = '-viridis',
      breaks = c(0, 1, 500, 1000, 5000, 10000, 50000, 2000000)
    ),
    fill.title = tm_legend(title = 'Population'),
    lwd = 0.1,
    border = 'white'
  ) +
  grid_final |> group_by(admin2Pcod) |> summarise(admin2Pcod = first(admin2Pcod)) |>
  tm_shape(name = 'Admin(2) boundaries') +
  tm_borders(lwd = 0.5) +
  tm_shape(markets_som_sf, name = "Market") +
  tm_dots()
```

### Traveltime to closest market

```{r, echo=F, warning=F}
tm_shape(grid_final, name = 'Traveltime to next market') +
  tm_polygons(
    'duration_h',
    fill.scale = tm_scale_intervals(
      values = 'hcl.heat',
      breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16)
    ),
    fill.title = tm_legend(title = 'Traveltime to closest market'),
    lwd = 0.1,
    border = 'white'
  ) +
  grid_final |> group_by(admin2Pcod) |> summarise(admin2Pcod = first(admin2Pcod)) |>
  tm_shape(name = 'Admin(2) boundaries') +
  tm_borders(lwd = 0.5) +
  tm_shape(markets_som_sf, name = "Market") +
  tm_dots()
```

### Closest market

```{r, echo=F, warning=F}
tm_shape(grid_final, name = 'Closest Market') +
  tm_polygons(
    fill = 'market_id',
    fill.scale = tm_scale_discrete(values = 'brewer.pastel1'),
    fill.title = tm_legend(title = 'Traveltime to closest market'),
    lwd = 0.1,
    border = 'white',
    fill.legend = tm_legend_hide()
  ) +
  grid_final |> group_by(admin2Pcod) |> summarise(admin2Pcod = first(admin2Pcod)) |>
  tm_shape(name = 'Admin(2) boundaries') +
  tm_borders(lwd = 0.5) +
  tm_shape(markets_som_sf, name = "Market") +
  tm_dots()
```

# Output

Grid export in geopackage format.

```{r, echo=F}
grid_final |> select(-c(snapped_centroid, centroid)) |>
  st_write('somalia_grid_access_markets.gpkg', append = F, quiet = T)
```

