---
title: "UKR primary health"
author: "Marcel Reinmuth"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message=FALSE, warning=FALSE)
```

This is an exploratory analysis for access to primary healthcare of places/settlements in Ukraine .

# Setup

## Libraries

The following libraries are used for in this R markdown document.

```{r "Libraries", warning = FALSE, message = FALSE}
library(sf)
library(tidyverse)
library(tmap)
library(units)

```

## input data

```{r}
ukr_health <- st_read("places_vinnytska.gpkg", quiet=T)
union_isochrones <- st_read("isochrones_union_vinnytska.gpkg", quiet=T)


ukr_health$travel_time <- ukr_health$travel_time |> set_units("seconds") |>  set_units("minutes") |> round(3)

ukr_health$travel_distance <- ukr_health$travel_distance |> set_units("meter") |> set_units("km") |> round(3)

ukr_health$min_distance <- ukr_health$min_distance |> set_units("meter") |> set_units("km") |> round(3)


union_isochrones$value <- union_isochrones$value |> set_units("seconds") |>  set_units("minutes") |> round(3)

union_isochrones <- union_isochrones |> arrange(desc(value))

```

# Explorative analysis

## Travel Time

```{r}
tmap_mode("view")

tm_basemap("OpenStreetMap") +
  tm_shape(ukr_health) +
  tm_symbols(
    col = "travel_time",
    palette = viridis::viridis(6),
    n=6,
    size = 0.1,
    border.lwd = 0,
    popup.vars = c("settlement", "travel_time", "travel_distance")
  )

```

## Travel distance vs. euclidean distance


### Distribution of distances

We see that the travel distances are much longer than the Euclidean distances.


```{r}

library(plyr)
library(ggplot2)


df <- rbind(
  data.frame(level4=ukr_health$level4,
             distance=ukr_health$min_distance,
             type="euclidean dist"),
  data.frame(level4=ukr_health$level4,
             distance=ukr_health$travel_distance,
             type="travel distance")
)

mu <- ddply(df, "type", summarise, grp.mean=mean(distance, na.rm=T))



ggplot(df, aes(x=distance, color=type, fill=type)) +
geom_histogram(aes(y=..density..), position="identity", alpha=0.5)+
geom_density(alpha=0.6)+
geom_vline(data=mu, aes(xintercept=grp.mean, color=type),
           linetype="dashed")+
labs(title="Distance histogram",x="Distance", y = "Density")

```



### Map


```{r}

m.1 <- tm_basemap("OpenStreetMap") +
  tm_shape(ukr_health) +
  tm_symbols(
    col = "travel_distance",
    palette = viridis::viridis(6),
    n=6,
    size = 0.1,
    border.lwd = 0,
    popup.vars = c("settlement", "travel_distance", "min_distance")
  )

m.2 <- tm_basemap("OpenStreetMap") +
  tm_shape(ukr_health) +
  tm_symbols(
    col = "min_distance",
    palette = viridis::viridis(6),
    n=6,
    size = 0.1,
    border.lwd = 0,
    popup.vars = c("settlement", "travel_distance", "min_distance")
  )

tmap_arrange(m.1,m.2, sync = T)

```



## Map Isochrones

```{r}

tm_basemap("OpenStreetMap") +
  tm_shape(union_isochrones |> st_make_valid()) +
  tm_polygons(
    col = "value",
    palette = viridis::viridis(6),
    n=6,
    alpha = 0.75,
    border.col = 0.1
  ) 
# +
#   tm_shape(ukr_health) +
#   tm_dots(
#     size = 0.1,
#     border.lwd = 0,
#     popup.vars = c("settlement", "travel_distance", "min_distance")
#   )
#   


```
